Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes

### 2020-04-01 treestats_paper.tex
cite ferretti; closes #42

\begin{figure}
    \begin{center}
    \includegraphics{figures/divergence_diagram_simple}
    \end{center}
    \caption{
        \textbf{FIXME:}
        An example of computing sequence divergence
        between red samples (triangles) and blue samples (squares)
        at a single site with a T$\to$G mutation, using the tree.
        The derived mutation is found in $a=2$ out of the $n_\text{red} = 3$ red samples
        and in $b=1$ of the $n_\text{blue}=2$ blue samples, and so the probability that a randomly chosen red node carries the derived allele
        but a randomly chosen blue node does not
        is given by the summary function, $f(a,b) = f(2, 1) = 2/6$.
        The complementary value, $f(1, 1) = 1/6$ is the probability that the mutation separates
        a randomly chosen red and blue pair, but with the blue node carrying the derived allele,
        so the total probability that a randomly chosen pair
        differs at this site is $f(2,1) + f(1,1) = 1/2$.
        \textbf{(a)} Weights are assigned to each subtree by adding together the weights of any children,
        plus the weight of the node itself, if it is a sample.
        \textbf{(b)} In this example, subtree weights $(a,b)$ record how many red samples ($a$)
        and how many blue samples ($b$) lie in that subtree.
        The summary function shown is applied to both the weight of the subtree
        and the remaining weight of each mutation,
        so for instance the mutation above the node with weights $(2,1)$
        contributes $f(2,1) + f(1,1) = 1/3 + 1/6 = 1/2$.
        \label{fig:divergence_diagram}
    }
\end{figure}

@@

% overview: general definition that is mostly functions of jAFS but includes others
In this paper we have described a general framework for summarizing genetic variation
and the underlying genealogies
that encompasses many standard population genetic statistics.
Many of these statistics are functions of the joint allele frequency spectrum,
but the framework is more general and can be used,
for example, to quantify associations between genotypes and phenotypes.
This generality greatly reduces the software development effort in
implementing statistics efficiently,
and it also allows users to easily explore new classes of statistics.
The range of statistics available to describe variation in a single exchangeable sample
(e.g., an isolated population) are well-understood,
but there are much larger and less well-explored classes of statistics
describing genetic variation between many populations
or across geographical space.
The statistics defined here are all additive along the genome:
if we have computed a statistic in equal-size windows along a chromosome,
then the value of the statistic for the entire chromosome is equal to
the average of the values in those windows.
Some commonly-used statistics (e.g., $F_{ST}$ or Tajima's $D$)
are not additive, but are ratios of additive statistics, so can be easily computed in this way.
% We have also defined site statistics to be additive across \emph{alleles},
% to allow the definitions to apply to sites with more than two alleles.
% This decision has some nice properties, but is not the only reasonable choice.
% Multilocus statistics that measure linkage disequilibrium or haplotype lengths
% do not fall in this category.
Extensions to statistics involving the pairwise joint distribution of genotypes across
sites~\citep{hudson2001twolocus} such as linkage disequilibrium are planned for future work.
Haplotype-based statistics may require different classes of algorithms.

>>>>

\begin{figure}
    \begin{center}
    \includegraphics{figures/divergence_diagram_simple}
    \end{center}
    \caption{
        An example of computing sequence divergence
        between red samples (triangles) and blue samples (squares)
        at a single site with a T$\to$G mutation, using the tree.
        The derived mutation is found in $a=2$ out of the $n_\text{red} = 3$ red samples
        and in $b=1$ of the $n_\text{blue}=2$ blue samples, and so the probability that a randomly chosen red node carries the derived allele
        but a randomly chosen blue node does not
        is given by the summary function, $f(a,b) = f(2, 1) = 2/6$.
        The complementary value, $f(1, 1) = 1/6$ is the probability that the mutation separates
        a randomly chosen red and blue pair, but with the blue node carrying the derived allele,
        so the total probability that a randomly chosen pair
        differs at this site is $f(2,1) + f(1,1) = 1/2$.
        \textbf{(a)} Weights are assigned to each subtree by adding together the weights of any children,
        plus the weight of the node itself, if it is a sample.
        \textbf{(b)} In this example, subtree weights $(a,b)$ record how many red samples ($a$)
        and how many blue samples ($b$) lie in that subtree.
        The summary function shown is applied to both the weight of the subtree
        and the remaining weight of each mutation,
        so for instance the mutation above the node with weights $(2,1)$
        contributes $f(2,1) + f(1,1) = 1/3 + 1/6 = 1/2$.
        \label{fig:divergence_diagram}
    }
\end{figure}

@@

% overview: general definition that is mostly functions of jAFS but includes others
In this paper we have described a general framework for summarizing genetic variation
and the underlying genealogies
that encompasses many standard population genetic statistics.
Many of these statistics are functions of the joint allele frequency spectrum,
but the framework is more general and can be used,
for example, to quantify associations between genotypes and phenotypes.
This generality greatly reduces the software development effort in
implementing statistics efficiently,
and it also allows users to easily explore new classes of statistics.
The range of statistics available to describe variation in a single exchangeable sample
(e.g., an isolated population) are well-understood
\citep{achaz2009frequency,ferretti2017decomposing},
but there are much larger and less well-explored classes of statistics
describing genetic variation between many populations
or across geographical space.
The statistics defined here are all additive along the genome:
if we have computed a statistic in equal-size windows along a chromosome,
then the value of the statistic for the entire chromosome is equal to
the average of the values in those windows.
Some commonly-used statistics (e.g., $F_{ST}$ or Tajima's $D$)
are not additive, but are ratios of additive statistics, so can be easily computed in this way.
% We have also defined site statistics to be additive across \emph{alleles},
% to allow the definitions to apply to sites with more than two alleles.
% This decision has some nice properties, but is not the only reasonable choice.
% Multilocus statistics that measure linkage disequilibrium or haplotype lengths
% do not fall in this category.
Extensions to statistics involving the pairwise joint distribution of genotypes across
sites~\citep{hudson2001twolocus} such as linkage disequilibrium are planned for future work.
Haplotype-based statistics may require different classes of algorithms.
